\pdftitle

### 2018-05-12 maxNt_v2.tex
V2

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-12-04T11:44:04+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[autostyle=false,autopunct=false,english=british]{csquotes}
\setquotestyle{american}

@@

% \DeclareMathSymbol{\theta}{\mathalpha}{egreeki}{"12}

@@

@@

%\newcommand*{\di}{\mathrm{d}}%differential

@@

\newcommand*{\mee}{maximum entropy}
\newcommand*{\me}{maximum-entropy}

%@@@@@@@@@@ new macros end @@@@@@@@@@

@@

\author{
  P.G.L. Porta Mana\\\texttt{mana@kth.org}
  \And
  V. Rostami\\\texttt{v.rostami@fz-juelich.de}
  \And
E. Torre\\\texttt{torre@ibk.baug.ethz.ch}
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

% LM: I'd like to avoid common beginnings like "Today blah blah is
% important" or "Blah blah is interesting/fundamental". They feel more like
% a mechanical ritual than a real tentative of connecting with the reader.
This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recordings.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-12T15:29:50+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[autostyle=false,autopunct=false,english=british]{csquotes}
\setquotestyle{british}

@@

 \DeclareMathSymbol{\theta}{\mathalpha}{egreeki}{"12}

@@

\pdfmapline{+dummy-space <dummy-space.pfb}\pdfinterwordspaceon%

@@

\newcommand*{\di}{\mathrm{d}}%differential

@@

\newcommand*{\mee}{maximum entropy}
\newcommand*{\me}{maximum-entropy}
\newcommand*{\ytheta}{\bm{\theta}}
%@@@@@@@@@@ new macros end @@@@@@@@@@

@@

\author{
  P.G.L. Porta Mana\\\texttt{piero.mana@ntnu.org}
  \And
  V. Rostami\\\texttt{v.rostami@fz-juelich.de}
  \And
E. Torre\\\texttt{torre@ibk.baug.ethz.ch}
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

% LM: I'd like to avoid common beginnings like "Today blah blah is
% important" or "Blah blah is interesting/fundamental". They feel more like
% a mechanical ritual than a real tentative of connecting with the reader.

Imagine that we have recorded the firing activity of a hundred neurons from
a particular brain area. This scenario is a concrete possibility thanks to
recent electrophysiological techniques \cite{berenyietal2014}. We bin and
binarize the recorded activity into sequences of $0$s (inactive) and $1$s
(firing) \cite{caianiello1961,caianiello1986}. We notice that groups of two
or more $1$s can appear in any time bin. We may ask several questions
looking at these groups. A recurring question is of this kind: can the
presence of groups of three or more simultaneous firings be
\enquote{explained} in terms of groups of just two or three? This kind of
question is very vague until we define what we mean with \enquote{explain}
or similar words.

A possible, precise definition is as follows. We want to predict, \ie\
assign a probability to, the population activity in a time bin, given the
population activities in the other bins:
\begin{equation}
  \label{eq:general_prediction}
  \p(\text{activity at one time bin} \|
  \text{activities at the remaining bins, and other info}).
\end{equation}
We can say that the simultaneous firing of at most two neurons
\enquote{explains} the activity if only the values of the first and second
empirical moments of the activities are relevant for the probability above:
\begin{multline}
  \label{eq:informal_sufficiency}
  \p(\text{activity at one time bin} \|
  \text{activities at the other bins, and other info}) ={}\\
  \p(\text{activity at some time bin} \|
  \text{1st \amp\ 2nd empirical moments, and other info});
\end{multline}
in statistical terms, the first and second moments are \emph{sufficient
  statistics}. Other definitions are possible, of course; the definition
above has the advantage of being mathematically very precise. In fact it
determines, from just the three basic laws of the probability calculus, the
mathematical form of the probability for the activity at any number of bins:
\begin{multline}
  \label{eq:sufficiency_maxent_family}
  \p(\text{activity at bins $1,2,\dotsc$} \|
  \text{other info}) ={}\\
  \shoveright{\int
    \pf(\text{activity at bin $1$} \|\ytheta)\,
    \pf(\text{activity at bin $2$} \|\ytheta)\dotsm
 \pf(\ytheta\| \text{other info})\,
  \di\ytheta,}
  \\[2\jot]
  \text{with \enquote{$\pf(\text{activity at time bin $i$} \|\ytheta)$}
    having maximum-entropy form}.
\end{multline}
The proof that the assumption of sufficient statistics leads to the
probability expression above is the Koopman-Pitman theorem
\cite{koopman1936,pitman1936,darmois1935} and can be found in
\citetext[\sect~4.5]{bernardoetal1994_r2000} and references therein. It can
be extended to higher moments and other statistics.

This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recordings.
